Peer-to-peer Browser Middleware for Communication and Consensus in an Unreliable Infrastructure

### 2015-12-07 research_class/workshop.tex
Add workshop.tex, paper-min.tex and update data files.

\title{Raft Distributed Consensus Algorithm over\\ WebRTC (Web Real-Time Communication)}

@@

\begin{abstract}
We describe the creation of a browser-based distributed peer-to-peer
application using WebRTC, PeerJS and Raft.js. WebRTC is a protocol
that modern browsers have begun to integrate which enables
browser-to-browser communication. PeerJS is a WebRTC client library
and extensible signaling server. Raft.js is an implementation in
JavaScript of the Raft distributed consensus protocol. These
technologies are combined to create a chat application that guarantees
a shared, consistent and linearized view of the chat history without
relying on a single centralized server for communication. We also
create a series of automated tests using the SlimerJS browser
automation tool and use them to show the correctness of the chat
application and also to measure the performance and scalability of the
application for various cluster sizes.
\end{abstract}

\section{Introduction}
In the past few years, browsers have quickly evolved from document
rendering engines into powerful platforms for building sophisticated
applications. Modern browsers are integrating new protocols for
communication including WebSockets for full-duplex browser-to-server
communication \cite{websockets:rfc6455} and WebRTC for direct
browser-to-browser communication \cite{webrtc:feb2015}. These new
capabilities enable software developers to treat web applications as
first class participants in distributed network applications.

These new browser capabilities also bring with them the traditional
challenges of building reliable distributed applications, such as:
concurrency, security, scalability, naming, discovery, replication,
relocation, etc. The solutions to many of these challenges involve
the proper management of distributed state. Distributed consensus is
one category of distributed state management in which multiple nodes
must come to agreement on the state of a single value or set of values. % TODO/CITE
The grandparent of distributed consensus algorithms is the Paxos
algorithm described by Leslie Lamport \cite{paxos:lamport98}. However,
Paxos has been notoriously difficult to understand and to implement
correctly \cite{paxos:chandra07}. Raft is a more recent distributed
consensus algorithm that aims for understandability and practicality
of implementation \cite{raft_thesis:ongaro14}.

In this paper we will describe an implementation of the Raft protocol
in a browser context using WebRTC as the communication channel. We
will describe the results of the implementation and look at some of
the unique challenges (and opportunities) we encountered with the
implementation. We will also explore some ideas for further study.

\section{Motivation/Use Cases}
The traditional use of distributed consensus is in the data-center
(private or cloud based), often as the core coordination component
of a replicated data-storage system. %TODO/CITE
In these examples, the set of nodes that compose the cluster tend to
be fairly static and changes to cluster membership are usually for the
purpose of replacing nodes. Changes to the long-term size of the
cluster are uncommon. %TODO/CITE

Most existing distributed browser applications are extensions of the
data-centric model in which browsers connect to load-balanced web
servers which are in close network proximity to the distributed
consensus cluster. %TODO/CITE
The backend web servers deliver the web application to the browser and
also serve as the conduit for all browser-to-browser communication and
state management.

However, requiring application messages and application state to
traverse the back-end servers is not always desirable for various
reasons, including the following:

\begin{enumerate}
\item \emph{security (privacy)}: application state and user messages
    may need to traverse directly from browser to browser rather than
    traversing a server.
\item \emph{locality}: the nature of the application may allow
    clusters of applications that are nearby on a network (e.g.
    intranet) to communicate and coordinate directly rather than
    passing all messages and state through a server back-end which may
    be considerably more remote.
\item \emph{bandwidth and scale}: a distributed browser based
    application may be able to avoid expensive bandwidth into and out
    of the backend data-center by passing messages and state directly
    between browsers.  Depending on the design of the application this
    may also enable better scalability then a more centralized model.
\end{enumerate}

%TODO: better lead in to this paragraph from above
Here are some specific use cases in which Raft over WebRTC may be
applicable:

\begin{enumerate}
\item \emph{Consistent order secure chat}: a chat room system where
    all users see the same message history but the messages never
    traverse a central server and there is no central arbiter
    of truth. The application code itself may still be delivered from
    a central web service. This use case was implemented
    for this paper.
\item \emph{Private multiuser document editing/whiteboard system}:
    Google Docs, among other examples, has demonstrated a powerful
    model for collaborative document creation. However, these systems
    all use a centralized servers to coordinate shared state and
    handle browser-to-browser communication. %TODO/CITE
    Using Raft over WebRTC could enable applications with similar
    functionality but without the central coordination and
    communication.
\item \emph{Browser based network games}: most network games have
    certain states that must be agreed upon by all nodes (e.g. avatar
    dead or alive, etc). This can be achieved with Raft over WebRTC
    without requiring any data to traverse a central server. In
    addition to reducing browser to server bandwidth requirements,
    this may also allow lower latency gaming if game instances are
    partitioned into clusters based on latency.
\end{enumerate}

@@

\subsubsection{Leader Election}

Each node of the Raft cluster maintains an ordinally increasing
current term value. When a Raft follower node does not receive
a heartbeat from a leader within a locally randomized election timeout
period, it transitions to candidate state, votes for itself, increases
its current term value by one and sends a requestVote RPC (with the
new term) to all the other nodes in the cluster. When a node receives
a requestVote RPC with a term that is greater than its own it will
become a follower (if not already), update its term to the new term,
and send a RPC response back that indicates a vote for that candidate.
However, a node may only vote for one candidate in a given term. If
a candidate receives votes from more than half of the cluster, then it
immediately becomes a leader and sends out a appendEntries RPC to
confirm leadership and reset the election timers on all the other
nodes \cite[Section~3.4]{raft_thesis:ongaro14}.

\subsubsection{Log Replication}

Each Raft node maintains a transaction log. The entries in the
transaction log each contain a term value in addition to the action to
apply to the state machine. Each node also keeps track of the most
recent log entry which is known to be committed (exists in the log of
more than half the nodes in the cluster).

When a new leader is elected, it begins sending appendEntries RPCs
to other nodes in the Raft cluster. These RPC messages contain
information about the state of its transaction log: index and term of
the latest entry, and the most recently committed entry. If the node
does not have an entry matching the most recent entry on the leader,
then it replies with false to indicate that its log is not up to
date. The next appendEntries RPC that the leader sends to that
node will contain the next oldest index and term. This will continue
until the leader discovers the latest log entry which is in agreement
(the node may not have any entries if it is new). Then the leader will
begin to propagate entries to the node in subsequent appendEntries
RPCs until the node is caught up. The leader continues sending
empty appendEntries as heartbeat RPCs to all the nodes until it
receives a new entry in its transaction log from a client
\cite[Section~3.5]{raft_thesis:ongaro14}.

\subsubsection{Safety}

In order to ensure that each state machine on every node executes
exactly the same commands in exactly the same order, the Raft system
provides safety guarantees. In particular, there are some restrictions
on whether a Raft node is actually eligible to become a leader, based on
the state of its transaction log compared to other node transaction
logs. A Raft node will only vote for a candidate if the candidate has
a log that is more up-to-date than the voter. A log entry with a later
term is always more up-to-date. If the log entries have the same term,
then the entry with a higher index is more up-to-date. In addition,
a Raft leader may not consider entries from a previous term to be
committed until it has committed at least one entry from its own
term
\cite[Section~3.6]{raft_thesis:ongaro14}.

Raft nodes must also persist certain properties to durable storage
before sending or receiving any RPCs. The properties that must be
persisted are: current term, current vote for this term (if any), and
either the full transaction log or the state machine plus any
unapplied transaction log entries
\cite[Section~3.8]{raft_thesis:ongaro14}.

\subsubsection{Membership Changes}

Raft leverages the replicated transaction log to accomplish live
cluster membership changes. Membership changes are accomplished by
adding or removing one Raft node at a time using special add/remove
log entries. As soon as the new entry is added to the log it becomes
effective on that node without waiting for the entry to be fully
committed across the cluster. However, a new cluster change entry may
not be added to the log until the previous one is committed. Once the
change entry is committed, the initiator of the change is notified,
removed servers can be shut down and another change entry may be added
to the log \cite[Section~4]{raft_thesis:ongaro14}.

\subsubsection{Log Compaction}

In order to keep the replicated transaction log from growing
indefinitely, Raft nodes should periodically compact their logs. There
are many different ways to accomplish log compaction and the best way
will depend on application requirements and on the specific nature of
the distributed state machine. For example, if the state machine is
simply a single shared counter, then any log entries before the most
recently committed entry may be safely discarded. The simplest and
most generic solution is often to provide some way of snapshotting the
entire state machine (including the term and log index it represents)
to disk at which point the all previously applied log entries can be
discarded. Snapshotting the state machine also means that the
implementation must provide a way for the leader to serialize and send
the current state machine to other nodes if their transaction log is
too old (e.g. when a new node is added)
\cite[Section~5]{raft_thesis:ongaro14}.

\subsubsection{Client Interaction}

Clients of the Raft cluster are able to interact with the distributed
state machine by sending RPCs to the leader in order to add command
entries to the transaction log. Clients first find the address of any
node in the cluster either via broadcast, via an external directory
service, or directly provided by the user. Once the client discovers
a node of the cluster, that node can either forward messages directly
to the leader node (if it is not the leader) or it can reject client
requests with a response indicating the address of the leader where
client requests should be redirected
\cite[Section~6.2]{raft_thesis:ongaro14}.

The Raft protocol can also provide strict linearizable semantics for
client commands/requests (reads and writes) for applications that
require it. In order to accomplish this, each client is given a unique
ID, and the client assigns a unique serial number to each command.
This prevents a single client command from being applied twice (leader
crash, network duplication) or lost (packet loss)
\cite[Section~6.3]{raft_thesis:ongaro14}.

Linearizability is important not just for writes/updates to the state
machine, but also for reads, which means that read requests must also
be entered as transactions in the log and committed to more than half
the Raft nodes before the leader can respond to the client. If reads
bypass the transaction log then those reads are serializable but not
necessarily linearizable. For example, a leader may have been deposed
and not realize it if the cluster is partitioned which could result in
reads of stale data (the leader of the larger partition may have
already committed new entries).
\cite[Section~6.4]{raft_thesis:ongaro14}.

\begin{figure*}[!t]
\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_architecture}}
\caption{Raft over RTC Architecture}
  \label{fig:raft_rtc_architecture}
\end{figure*}

\subsection{Raft.js}

Raft.js is an implementation of the Raft algorithm in JavaScript that
was created by the author of this paper \cite{raft.js}.  Raft.js is
designed to run in either a browser environment or within Node.js
(server-side JavaScript). Raft.js implements the Raft algorithm as
described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}.

\subsubsection{Modular Design}

The implementation of the Raft algorithm is in the
\texttt{RaftServerBase} class (in \texttt{base.js}). The base class
omits the implementation of functions with side-effects such as RPC
communication, durable storage, scheduling/timeouts, or state machine
command/management. In order to create a working implementation these
functions must be provided either as configuration parameters when the
class is instantiated or by sub-classing the class.

The modular design of the Raft.js implementation allows it to be
easily used in different contexts. For example, the \texttt{RaftServerLocal}
class (\texttt{local.js}) implements RPC calls using plain JavaScript function
calls between different instances of the class (nodes) in the same
execution context. For example, the online Raft visualization at
\href{http://kanaka.github.io/raft.js/}{http://kanaka.github.io/raft.js/}
\cite{raft.js:visualization} uses the \texttt{RaftServerLocal} class
and instantiates it using a scheduling function that executes one
scheduled event each time the user clicks on the ``Step" button rather
than due to the passage of time.

The \texttt{RaftServerHttp} class (\texttt{http.js}) embeds a simple
web server into each Raft node and then uses normal HTTP requests to
send RPC messages between Raft nodes.

For this paper, the \texttt{RaftServerLocal} class was extended with
the capability of sending RPCs over the WebRTC Data Channel.

\subsubsection{Differences between Raft and Raft.js}

The current version of Raft.js implements most of the Raft algorithm
including: leader election, log replication, safety, membership
changes, basic client interaction, and read-only operation
optimizations. % TODO: read-only operations needs mentioning earlier
Log compaction and full client linearizability are not currently
implemented (clients and client transactions are not assigned unique
IDs). Log compaction is important for long-running production systems
in order to limit memory growth and allow new nodes to catch up their
logs more efficiently. However log compaction is not required for
verifying overall system correctness and does not significantly impact
short-running test configurations. Full client linearizability is
an important feature for simplifying certain types of distributed
applications, however the implementation of linearizability requires
some significant additions to the Raft implementation (e.g. assigning
client IDs and sequence IDs and tracking timed-out clients). For the
chat application, simply adding a single atomic check-and-set state
machine command is sufficient to achieve a linearized chat history.
This does impose a slight additional cost on the client because it
must retry the atomic operations until they succeed. The author
intends to implement both log compaction and full client
linearizability in the future, however these do not have a significant
impact on the ability to test Raft combined with WebRTC.

The Raft protocol has undergone some revisions since the original
draft paper \cite{raft_paper:ongaro14} (including incorporation of
suggestions by the author of this paper). These changes include
cluster membership simplifications, separating the concepts of
committing entries to the log and applying entries to the state
machine, clearer client interaction, etc. As part of the work for
implementing Raft over WebRTC, the Raft.js implementation was updated
to reflect these changes to the Raft protocol definition.

The are also some other differences between Raft.js and Raft as
described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}:

\begin{enumerate}
\item The paper describes the Raft RPCs in terms of synchronous stream
    based request/response model where requests and responses are
    automatically correlated together by a lower level of the software
    stack. For example with HTTP 1.0 \cite{http:rfc1945} client
    requests are followed by a server response on any given TCP
    stream. Most remote procedure call mechanisms also use
    a synchronous request/response model. %TODO/CITE
    WebRTC DataChannel connections are message based rather than
    request/response stream based. %TODO/CITE
    In the original Raft definition of RPC, request and response messages
    almost contain enough information to function correctly in an
    environment without builtin request/response correlation, such as
    UDP or SCTP when used with reliable transport disabled. %TODO/CITE
    Raft.js includes a small modification to the response RPCs for
    requestVote and appendEntries to add a `sourceId' field that
    enables the protocol to function with message based transports.
\item The Raft paper is clear that each membership change must be
    delayed until all the previous membership change is committed
    (sequential) \cite[Section~4.1]{raft_thesis:ongaro14}. It is
    implied that the leader node should track all pending membership
    change requests and apply them one at a time.  However, in
    Raft.js, if a client requests a membership change while another
    change request is pending (uncommitted), the leader will reject
    the change request with a status of `PENDING\_CONFIG\_CHANGE'.
    This keeps the core Raft implementation simpler with the trade-off
    being that the client implementation of membership change requests
    is slightly more complex due to retry logic.
\end{enumerate}

@@

\subsubsection{WebRTC Discovery and Signaling}

One of the challenges with browser-to-browser communication (and
direct peer-to-peer communication in general) is that the web browser
environment or user agent (UA) is usually running behind a firewall in
a private internet subnet. This complicates direct communication in
several ways: the public Internet address of the user agent is unknown
and dynamic, inbound connections are denied by default, and outbound
connections are network address translated (NATed)
\cite[Section~3.4]{draft-rtcweb-transports}.

To establish a direct communication channel between browsers, WebRTC
uses a mechanism called Interactive Connectivity Establishment (ICE)
that is similar to the Session Initiation Protocol (SIP) commonly used
for Voice-over-IP (VoIP) communication \cite{RFC5245:ice}. This
involves a third party server known as a signaling server. Each user
agent (browser) connects to this server at a well known address in
order to establish a direct connection to other user agents. The
signaling server is also used to communicate session control messages
(open, close, error) and to negotiate media capabilities. %TODO/CITE
The signaling server may optionally provide other services such as
higher level session management, HTTP serving of the web application,
address book service, client presence/availability, etc. %TODO/CITE

@@

There are three primary WebRTC interfaces:

\begin{enumerate}
\item \emph{RTCPeerConnection} \cite{webrtc:rtcpeerconnection:feb2015}:
    The RTCPeerConnection API represents a direct connection between
    two browsers. Each RTCPeerConnection object has an associated
    Interactive Connectivity Establishment (ICE) agent that is used to
    establish network connectivity through firewalls and NAT services.

\item \emph{MediaStream~(aka~getUserMedia)} \cite{webrtc:media-stream:feb2015}:
    The MediaStream API represents a stream of audio or video data.
    MediaStream data can easily be sent and received over
    RTCPeerConnections. The MediaStream API definition provides
    a mechanism that allows JavaScript to enumerate the native
    MediaStreams that are provided by the browser such as a local web
    camera or a live video stream of the desktop (for desktop
    sharing).  Although support for MediaStreams was the original
    impetus for the creation of the WebRTC APIs and protocols, the
    Raft over WebRTC implementation does not make use of the
    MediaStream API.

\item
    \emph{RTCDataChannel} \cite{webrtc:datachannel:feb2015}:
    The RTCDataChannel API provides a mechanism for sending and
    receiving generic data between User Agents (browsers). A single
    RTCPeerConnection can contain multiple RTCDataChannels. The
    RTCDataChannel API is modeled after the WebSocket API, however,
    unlike the WebSocket API which is reliable and guaranteed order
    transport (like TCP), each RTCDataChannel can be configured with
    different reliability and ordering settings. For example, by
    relaxing the reliability and ordering constraints on
    a RTCDataChannel, the average latency and jitter per message can
    be decreased even though some messages may be dropped or arrive
    out of order. The Raft protocol is specifically designed to
    support dropped or out-of-order messages so the transport options
    available with RTCDataChannel can be used to gain extra
    efficiency.
    %TODO/CITE: more of them above
\end{enumerate}

\subsubsection{WebRTC Protocol}

There is a large suite of protocols that are required for a working
WebRTC system. The Internet Engineering Task Force (IETF) organization
is responsible for specifying and/or standardizing (where necessary)
the protocols that are part of WebRTC \cite{draft-rtcweb-overview}
\cite{draft-rtcweb-security-arch} \cite{draft-rtcweb-security}. In many
cases an existing protocol is applicable and is referenced or
extended as part of the WebRTC suite. %TODO/CITE

The protocols that make up the WebRTC suite are categorized into
five general functionality groups:

\begin{enumerate}
    \item data transport
        \ifdefined\OPTIONAL
        \\
        \cite{draft-rtcweb-transports}
        \cite{draft-rtcweb-qos}
        \cite{RFC4960:sctp}
        \cite{RFC3758:sctp-reliability}
        \cite{draft-sctp-dtls-encaps}
        \cite{draft-mmusic-sctp-sdp}
        \cite{draft-sctp-ndata}
        \cite{draft-rtcweb-rtp-usage}
        \cite{draft-rtcweb-data-channel}
        \cite{draft-rtcweb-data-protocol}
        \cite{draft-rtcweb-alpn}
        \cite{RFC5245:ice}
        \cite{RFC5128:p2p-nat}
        \cite{RFC5389:stun}
        \cite{RFC3489:stun-udp}
        \cite{RFC5766:turn}
        \cite{RFC6156:turn-ipv6}
        \cite{RFC6544:tcp-ice}
        \cite{RFC4571:rtp-rtcp}
        \cite{RFC5764:dtls-srtp}
        \fi
    \item data framing and security
        \ifdefined\OPTIONAL
        \\
        \cite{RFC3550:rtp}
        \cite{RFC3711:srtp}
        \cite{draft-rtcweb-rtp-usage}
        \cite{draft-rtcweb-data-channel}
        \cite{draft-rtcweb-data-protocol}
        \fi
    \item connection management
        \ifdefined\OPTIONAL
        \\
        \cite{RFC2327:sdp}
        \cite{RFC3264:sdp-offer-answer}
        \cite{draft-rtcweb-jsep}
        \cite{RFC5245:ice}
        \cite{draft-mmusic-trickle-ice}
        \cite{RFC5763:srtp-using-dtls}
        \fi
    \item data formats
    \item presentation and control
    \item local system support functions
\end{enumerate}

The last three functionality groups focus on interactive audio and video
applications and are less relevant to Raft over WebRTC. \ifdefined\OPTIONAL
Some of the important standards for the first three functionality
groups are cited above. \fi

The signaling transport itself is not defined as part of WebRTC.
This is up to the application designer. The JavaScript interface used
for establishing the initial connection (signaling) in the WebRTC 1.0
standard is somewhat complicated. A competing standard called Object
Real-time Communication (ORTC) is working to define a simpler and more
modular JavaScript API for WebRTC signaling \cite{ortc2014}. However,
at the time of this paper, the ORTC interface is not yet widely
supported and is still in a state of flux. The use of PeerJS provides an
abstraction that is likely to transparently support ORTC in the future.

\ifdefined\OPTIONAL

@@

\subsubsection{PeerJS Server}

The PeerJS project provides a simple WebRTC signaling server called
PeerServer. This server uses HTTP and WebSockets transport protocols
to perform WebRTC signaling on behalf of browser clients. The
PeerServer server is also extensible so that more advanced services
can be built with it. The PeerJS project includes an extension to
PeerServer called ExpressPeerServer that combines a PeerJS signaling
server with the popular Node.js ExpressJS web application framework.
% (TODO/CITE)

\subsubsection{PeerJS Client Library}

The WebRTC API standards are not yet
finalized and so different browsers and browser releases may support
different versions of the WebRTC draft APIs. %TODO/CITE
The PeerJS client library abstracts over these differences and
provides a common interface so that the application developer does not
have to deal directly with browser differences.

The PeerJS client library also provides a default signaling mechanism
that is designed to operate with a PeerJS signaling server. This is
particularly helpful because much of the WebRTC signaling transport
and protocol is not defined as part of the WebRTC standardization
effort. %TODO/CITE
The PeerJS organization provides a cloud-hosted version of a PeerJS
compatible signaling server, however application developers can also
run their own signaling service using PeerJS PeerServer.

\fi  % \ifdefined\OPTIONAL

%\begin{figure*}[!t]
\begin{figure}[!t]
%\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_sequence}}
\centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_sequence}}
\caption{Raft over RTC Sequence Diagram}
  \label{fig:raft_rtc_sequence}
\end{figure}

\section {Design and Implementation}

\ifdefined\OPTIONAL

\subsection{Incremental Steps}

During the planning phase, the project was broken down into ten incremental steps:

\begin{enumerate}
\item Simple WebRTC messaging: implement a small JavaScript
    application that uses the PeerJS client library and an unmodified
    PeerJS signaling server. Demonstrate communication over WebRTC
    DataChannel using two browser instances (or browser tabs).
\item Raft.js over WebRTC: combine the existing Raft.js implementation
    with the small JavaScript application above to demonstrate Raft
    over WebRTC. This step uses a statically configured three node
    cluster (three browser tabs). The PeerJS signaling server is
    polled to retrieve the list of other clients and the Raft cluster
    is started when the number of PeerJS clients reaches three.
\item Update Raft.js protocol: update the Raft.js implementation to
    match the latest description of the Raft protocol
    \cite{raft_thesis:ongaro14}. This includes the implementation of
    the simplified membership change algorithm (sequential changes)
    and the separation of the log commit and state machine apply
    concepts.  Additionally, the new Raft.js implementation always
    starts a new Raft cluster as a cluster with a single node and
    additional nodes are dynamically added to the cluster
    \cite[Section~4.4]{raft_thesis:ongaro14}.
\item Dynamic Raft.js over WebRTC: enhance the signaling server with
    support for multiple sessions/channels to enable separate Raft
    clusters with the same signaling server. Start each cluster with
    a single node and automatically add and remove Raft nodes as they
    come and go.
\item Visual representation: display Raft node state and RPC
    statistics in addition to the internal Raft logging messages.
\item Sessionless/Datagram RPC messages: change Raft RPC requests and
    response messages so that RPC responses have enough context to be
    handled separately from RPC requests. Specifically this enables
    Raft to support datagram (sessionless) network transports like
    WebRTC.
\item Atomic state machine operation: implement a state machine
    update operation that only succeeds if no other operation has been
    applied to the same key since the last time this client read from
    that key.
\item Automatic client request forwarding: extend the client request
    RPC to also be sessionless/datagram based. Implement an
    asynchronous JavaScript function that determines which node is the
    leader and forwards client request RPCs to the leader. This
    function must also correlate RPC responses to the client request
    and invoke the callback for the original request.
\item Chat application: add the ability to send a chat message to the
    log using the client request forwarding functionality together
    with atomic state machine operation. Show the current chat history
    based on the chat entries in the state machine.
\item Server as Raft.js peer (aspirational): enable server nodes
    (Raft.js running on Node.js) to fully participate as Raft cluster
    nodes.
\end{enumerate}

Steps 1-9 were successfully implemented for this paper. Step 10 was an
aspirational goal because this functionality is not yet supported in
PeerJS \cite{peerjs:103}.

\fi  % \ifdefined\OPTIONAL

\subsection{Server}

The file `rtc\_server.js' implements the signaling server for the Raft
over WebRTC application. The server extends the PeerJS
`ExpressPeerServer' object in order to provide WebRTC signaling. In
addition to the signaling function, the server also provides the
following service endpoints:

\begin{enumerate}
\item Static web service endpoint: this serves the static HTML, CSS
    and JavaScript that make up the Raft over WebRTC web application.
\item New channel endpoint: when a browser loads this endpoint, a new
    PeerJS channel is created and the browser is redirected to a URL
    for loading the main application (via static file service above).
    The redirect URL has a query string appended that contains the
    channel ID of the new channel that was just created. In addition
    the redirect URL has a fragment identifier (hash/history) that
    communicates to the web application that this is the first server
    in the channel.
\item Peer list endpoint: this endpoint takes a channel ID and returns
    the list of WebRTC peers (browser user agents) that are currently
    connected to that channel.

\end{enumerate}

Refer to Section~\ref{section:design-client} and
Figure~\ref{fig:raft_rtc_sequence} for a description of how these
endpoints interact.

\subsection{Client}
    \label{section:design-client}

The Raft over WebRTC testing application is designed as a single web
page (\texttt{rtc.html} or \texttt{chat.html}). The application has
different starting behavior depending on the value of the URL query
string (the URL component following the ``?" and up to the ``\#") and
the fragment identifier (the URL component after the ``\#").

When a new Raft cluster is created, the user agent (browser) connects
to the server's new channel endpoint
(Figure~\ref{fig:raft_rtc_sequence}~[1]). This endpoint causes the
server to create a new channel for PeerJS signaling
(Figure~\ref{fig:raft_rtc_sequence}~[2]). Once this is complete the
server responds to the browser with a redirect message
that contains the new channel identifier in the query parameter and an
indication that this is the very first Raft cluster node in the
fragment identifier (Figure~\ref{fig:raft_rtc_sequence}~[3]).

The browser follows the redirect and loads the static web application
at the static file endpoint provided by the server
(Figure~\ref{fig:raft_rtc_sequence}~[4]). This includes the
main application page (\texttt{rtc.html} or \texttt{chat.html}), the
core raft algorithm (\texttt{base.js} and \texttt{local.js}), the
PeerJS client library (\texttt{peer.js}), the actual Raft over WebRTC
implementation (\texttt{rtc.js}), a layout stylesheet
(\texttt{web/demo.css}), and higher-level application code if that is
being loaded (\texttt{chat.js})
(Figure~\ref{fig:raft_rtc_sequence}~[5]).

If the URL fragment identifier is set to ``firstServer", then this Raft
node is initialized as the leader of a single node cluster and all
normal timers are started. The fragment identifier is then unset. The
query parameter contains the ID of this cluster (the PeerJS channel).

Subsequent nodes are added to the same Raft cluster by starting a new
browser context (separate browser, browser window, or browser tab) and
loading the same URL without the ``firstServer" fragment identifier
(Figure~\ref{fig:raft_rtc_sequence}~[9]). There is a convenience link
provided in the application page that opens a new browser window with
a new cluster node (see \ref{fig:browser_1}).

When each page first loads, a PeerJS signaling connection is
established with the server
(Figure~\ref{fig:raft_rtc_sequence}~[6]~and~[11]). Once this
connection is established, the application will be notified when any
PeerJS clients connect to or disconnect from the server.  After the
PeerJS signaling connection is established, the client immediately
queries the Peer list endpoint
(Figure~\ref{fig:raft_rtc_sequence}~[7]~and~[12]~and~[14]]) to discover
any clients that were connected prior to when this node connected to
the server (Figure~\ref{fig:raft_rtc_sequence}~[8]~and~[13]~and~[15]]).

The web application also begins a periodic async polling function
(\texttt{addRemoveServersAsync}) when the page loads. When the node is a Raft
leader, the function compares the current active list of PeerJS clients to
the Raft cluster membership list. If there are any differences between
the two lists, then any new clients are added to the Raft cluster and
missing clients are removed from the Raft cluster one at a time using
the \texttt{addServer} and \texttt{removeServer} RPC calls. Servers
additions are prioritized before removals 
in order to maximize availability
\cite[Section~4.4]{raft_thesis:ongaro14}.

@@

In addition to the low level Raft message log, the following
statistics are shown in the application:

>>>>

\title{Peer-to-peer Browser Middleware for Communication and Consensus in an Unreliable Infrastructure}

@@

\begin{abstract}
We describe a browser-based peer-to-peer computing system/middleware that enables distributed consensus that is tolerant of degraded and/or unreliable centralized infrastructure. The system is composed of a JavaScript implementation of the Raft distributed consensus protocol (Raft.js) together with the WebRTC browser-to-browser communication protocol. The Raft over WebRTC middleware is used to build a demonstration chat application that guarantees a shared, consistent and linearized view of the chat history without relying on a single centralized server for communication. The correctness, robustness, performance, and scalability of the system is verified with a series of automated tests implemented using the SlimerJS browser automation tool.
\end{abstract}

\section{Use Case/Scenario}
Consider a scenario where emergency responders and volunteers are responding to a city-scale disaster in which utilities such as electrical power and communication systems are highly degraded or completely unavailable for an extended period. Emergency management officials have divided the city into zones each of which is assigned several dozen volunteers who systematically search the neighborhoods for people that are injured or in need and also for damaged infrastructure that needs to be repaired.

Responders and volunteers have access to a distributed disaster response datastore that is accessed and updated via a browser based application. Volunteers use their own mobile device and only require a browser with WebRTC capability to run the application. The application communicates over the cellular network (existing or provided by mobile emergency response cell stations) or mesh network (although this may also require additional installed software on mobile devices in order to support it). Once loaded and running, the application does not require any centralized servers or database system. The application uses the Raft distributed consensus protocol to maintain a distributed but consistent view of the datastore among all participating client nodes. Also, the distributed cluster is dynamic and allows nodes to join and leave over time without losing state or synchronization of the datastore.

The above capabilities are demonstrated using a distributed chat application implemented with the Raft over WebRTC middleware. This chat application uses direct browser-to-browser communication over WebRTC and maintains a consistent view of the chat history (exact same messages and ordering is seen by all nodes) and is fault tolerant (almost half the cluster nodes can fail simultaneously and the application will remain available without dataloss for the remaining clients).

@@

Raft leverages the replicated transaction log to accomplish live cluster membership changes. Membership changes are accomplished by adding or removing one Raft node at a time using special add/remove log entries. In order to keep the replicated transaction log from growing indefinitely, Raft nodes should periodically compact their logs. The simplest and most generic solution is often to provide some way of snapshotting the entire state machine (including the term and log index it represents) to disk at which point the all previously applied log entries can be discarded. Clients of the Raft cluster are able to interact with the distributed state machine by sending RPCs to the leader in order to add command entries to the transaction log. Clients first find the address of any node in the cluster and are then redirected to communicate directly with the leader. The Raft protocol also provides strict linearizable semantics for client commands/requests (reads and writes) for applications that require it.

@@

One of the challenges with browser-to-browser communication (and direct peer-to-peer communication in general) is that the browser is usually running behind a firewall in a private internet subnet. To establish a direct communication channel between browsers, WebRTC uses a mechanism called Interactive Connectivity Establishment (ICE) that is similar to the Session Initiation Protocol (SIP) commonly used for Voice-over-IP (VoIP) communication \cite{RFC5245:ice}. This process requires out-of-band exchange of information usually involving a third party server known as a signaling server. The signaling server may optionally provide other services such as higher level session management, HTTP serving of the web application, address book service, client presence/availability, etc.

@@

There are three primary WebRTC interfaces: \emph{RTCPeerConnection} \cite{webrtc:rtcpeerconnection:feb2015} for establishing direct connections between browsers, \emph{RTCDataChannel} \cite{webrtc:datachannel:feb2015} for sending and receiving generic data between browsers, and \emph{MediaStream~(aka~getUserMedia)} for managing audio and video streams (not relevant to Raft over WebRTC). The RTCDataChannel API supports both multiple modes for reliability and ordering which allows average latency/jitter to be decreased at the cost of some messages being dropped or arriving out of order. The Raft protocol is specifically designed to support dropped or out-of-order messages so the transport options available with RTCDataChannel can be used to gain extra efficiency.

\subsubsection{WebRTC Protocols}

There is a large suite of protocols that are required for a working WebRTC system. The Internet Engineering Task Force (IETF) organization is responsible for specifying and/or standardizing (where necessary) the protocols that are part of WebRTC \cite{draft-rtcweb-overview} \cite{draft-rtcweb-security-arch} \cite{draft-rtcweb-security}. In many cases an existing protocol is applicable and is referenced or extended as part of the WebRTC suite. The protocols that make up the WebRTC suite are categorized into five general functionality groups: data transport, data framing and security, connection management, data formats, presentation and control, and local system support functions. The last three groups focus on interactive audio and video applications and are not relevant to Raft over WebRTC.

\section{Design and Implementation}

\begin{figure}[!t]
  \centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_architecture}}
  \caption{Raft over RTC Architecture}
    \label{fig:raft_rtc_architecture}
\end{figure}

\subsection{Raft.js Core Library}

Raft.js is an implementation of the Raft algorithm in JavaScript that was created by the author of this paper \cite{raft.js}.  Raft.js is designed to run in either a browser environment or within Node.js (server-side JavaScript). Raft.js implements the Raft algorithm as described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}. The core Raft algorithm is defined in the \texttt{RaftServerBase} abstract class (in \texttt{base.js}) which omits the implementation of methods with side-effects such as RPC communication, durable storage, scheduling/timeouts, or state machine command/management. The full instantiation of the class (in \texttt{rtc.js} and \texttt{chat.js}) adds state machine commands and the ability to send RPCs over the WebRTC Data Channel.

@@

\subsection{Signalling Server}

The file `rtc\_server.js' implements the signaling server for the Raft over WebRTC application. The server extends the PeerJS `ExpressPeerServer' object in order to provide WebRTC signaling. In addition to the WebRTC signaling function the server also serves static web resources (HTML, CSS, and JavaScript) and WebRTC channel management endpoints.

\subsection{Client}

The Raft over WebRTC chat application is a single web page (\texttt{chat.html}). When a new Raft cluster is created, the browser connects to the signalling server's new channel endpoint (Figure~\ref{fig:raft_rtc_sequence}~[1]). This endpoint causes the server to create a new channel for PeerJS signaling (Figure~\ref{fig:raft_rtc_sequence}~[2]). Once this is complete the server responds to the browser with a redirect message that contains the new channel identifier in the query parameter and ``firstServer” in the fragment identifier to indicate that this is the very first Raft cluster node (Figure~\ref{fig:raft_rtc_sequence}~[3]).

Subsequent nodes are added to the same Raft cluster by starting a new browser context (separate browser, browser window, or browser tab) and loading the same URL without the ``firstServer" fragment identifier (Figure~\ref{fig:raft_rtc_sequence}~[9]). There is a convenience link provided in the application page that opens a new browser window with a new cluster node.

When each page first loads, a PeerJS signaling connection is established with the server (Figure~\ref{fig:raft_rtc_sequence}~[6]~and~[11]). The web application also begins a periodic async polling function (\texttt{addRemoveServersAsync}) that compares the current active list of PeerJS clients to the Raft cluster membership list. If there are any differences between the two lists, then any new clients are added to the Raft cluster and missing clients are removed from the Raft cluster one at a time using the \texttt{addServer} and \texttt{removeServer} RPC calls. Servers additions are prioritized before removals in order to maximize availability \cite[Section~4.4]{raft_thesis:ongaro14}.

@@
